---
title: "Inspecting and Debugging Submissions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspecting and Debugging Submissions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is for diagnosing a job that does not submit, stays queued, or
does not run as expected. Start with `vignette("introduction", package =
"hpcR")` for the normal build, submit, and tracking workflow.

The examples below create and compile a job but do not submit work to a local
shell or cluster scheduler when the vignette is built.

## Build a Job to Inspect

Create a small job specification. Slurm is used here because compilation can
prepare its scheduler directives without requiring access to a Slurm cluster.

```{r}
library(hpcR)

job_dir <- tempfile("hpcr-debug-")
dir.create(job_dir)

script_path <- file.path(job_dir, "analysis.R")
writeLines("print('analysis complete')", script_path)

debug_job <- rjob("debug_analysis") +
  script(script_path) +
  job_directory(job_dir) +
  scheduler("slurm") +
  resources(
    n_nodes = 1,
    n_cores = 2,
    wall_time = format_wall_time(minutes = 10),
    total_memory = "2G"
  ) +
  packages("stats")

summary(debug_job)
```

`summary()` is the first check: it confirms the script, working directory,
scheduler, and resource requests recorded in the job object. It describes the
job configuration; it is not a rendered scheduler command.

## Preflight Checks

Check that the submission host can see the script and working directory, and
that the requested R packages are installed there.

```{r}
required_packages <- "stats"

c(
  script_exists = file.exists(script_path),
  job_directory_exists = dir.exists(job_dir),
  rscript_available = nzchar(Sys.which("Rscript")),
  packages_available = all(vapply(
    required_packages,
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
)
```

For a cluster job, the script and working directory must also be visible on the
compute node. Packages must be installed in library paths visible to the R
process on that node, which can differ from the submission host.

## Package and Library-Path Setup

`packages()` is a declaration and loading interface. It records packages that
the run wrapper must find and attach. At submission, `hpcR` checks the current
library search path and exports it as `R_LIBS` to the compute process. If one
declared package is missing, the job normally fails before sourcing the user
script.

The run wrapper never installs packages. You can instead request installation
in the submission session with `install = "ask"` (interactive use) or
`install = "always"` (non-interactive use). An installation targets one
writable library, using `R_LIBS_USER` by default or `install_library` when
specified. Make sure that library is mounted and readable on every compute node
that can run the job.

For example, this job asks before installing missing packages in a project
library:

```{r, eval = FALSE}
debug_job <- debug_job +
  packages(
    c("data.table", "stats"),
    install = "ask",
    install_library = "/shared/projects/example/R/library"
  )
```

Use `libraries()` to augment or prioritize the job's effective R library
search path. This is useful for a site library, for a script that consistently
uses `pkg::function()` without attaching packages, or for setting a
cluster-specific user library. Explicit `R_LIBS` paths are prepended to the
rest of the effective library search path; `R_LIBS_USER` is passed through as
the R-user-library setting.

```{r, eval = FALSE}
debug_job <- debug_job +
  libraries(
    R_LIBS = "/shared/site/R/library",
    R_LIBS_USER = "~/R/library"
  )
```

`libraries()` accepts paths without checking that they exist on the
submission host, because some sites expose compute-node-specific paths. Verify
those paths in a short test job when they are not shared with the submission
host. Installation needs a path that is writable from the submission session.

Confirm that the scheduler and accounting commands required by your backend are
available on the host where you submit:

```{r, eval = FALSE}
Sys.which(c("sbatch", "sacct"))       # Slurm
Sys.which(c("qsub", "qstat", "tracejob")) # TORQUE
```

## Inspect Compiled Metadata

`compile_job()` prepares the bundled submission wrapper, scheduler arguments,
and runtime environment. It does not submit the job or contact the scheduler.

```{r}
compiled_job <- compile_job(debug_job)
```

The compiled metadata is stored in an internal property. It is useful for
interactive diagnosis, but scripts should not depend on its exact layout; a
future public submission-plan interface may replace it.

```{r}
compiled_metadata <- S7::props(compiled_job@.compiled)

compiled_metadata$submit_control$scheduler_arguments
compiled_metadata$submit_system_file
compiled_metadata$env_variables[c(
  "job_dir", "input", "scheduler_name", "packages", "R_LIBS", "R_LIBS_USER"
)]
```

For this Slurm job, the scheduler arguments should include the job name, node
count, core count, wall time, and total memory. The submission system file is
the package-supplied Slurm wrapper, and the selected environment values show
the script, working directory, package declaration, and library paths it will
receive. `R_LIBS_USER` is omitted when it was not explicitly configured.

`summary(compiled_job)` deliberately remains a concise configuration summary.
It does not print compiled metadata, so use the diagnostic view above only when
you need to investigate how a job was prepared.

## Submit and Track

Once the preflight checks look correct, submit the configured job. This code is
not run while the vignette is built because it contacts a real scheduler.

```{r, eval = FALSE}
job_id <- debug_job |> submit()
job_id
```

Use the returned ID to inspect status. `hpcR` standardizes scheduler states
such as `queued`, `running`, `complete`, `failed`, and `cancelled`.

```{r, eval = FALSE}
check_job_status(
  job_ids = job_id,
  scheduler_name = "slurm"
)

wait_for_job(
  job_ids = job_id,
  scheduler_name = "slurm",
  repolling_interval = 60,
  max_wait = 60 * 60
)
```

`wait_for_job()` returns `TRUE` when all jobs complete successfully and `FALSE`
when a tracked job fails or is cancelled. It stops with an error on timeout by
default.

## Diagnose Common Failures

If submission fails before an ID is returned, confirm that the scheduler
command is available, that your account is authorized to submit, and that your
cluster's required settings are present. `rjob()` currently translates job
name and resource properties into scheduler directives; site-specific options
such as account, partition, and QoS are not job properties.

If a job stays queued, inspect the scheduler's reason for waiting. Queue policy,
available resources, requested wall time, and account or partition rules are
managed by the cluster rather than by `hpcR`.

If a job fails after it starts, inspect the scheduler output and error files
using the returned job ID. Common causes are an inaccessible script or working
directory, a package missing on the compute node, or an error raised by the R
script itself. For a missing package, confirm that it was provisioned before
submission and that its library path is visible on the compute node. The
`hpcR` run wrapper echoes sourced R code to standard output, which can help
identify the failing expression.

For an R-only check before using a scheduler, return to the introduction
vignette and run the local backend example. It separates R-script and package
problems from cluster configuration problems.
