---
title: "Introduction to hpcR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hpcR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`hpcR` provides a user-friendly interface for preparing R jobs, submitting them to a
cluster scheduler, and checking on them after submission. The core workflow is:

1. create an R job with `rjob()`;
2. add job properties with `+`;
3. inspect the job with `summary()`;
4. submit the job with `submit()`;
5. track it with `check_job_status()` or `wait_for_job()`.

This vignette walks through that workflow with a minimal R script. The examples
that build and inspect job objects are evaluated when the vignette is built.
Examples that would submit work to Slurm, TORQUE, or a local shell are shown but
not run.

## Create a Script

An `hpcR` job usually points to a script that already exists on disk. Here we
create a small script in a temporary directory so that the rest of the vignette
has a concrete file to work with.

```{r}
library(hpcR)

job_dir <- tempfile("hpcr-demo-")
dir.create(job_dir)

script_path <- file.path(job_dir, "summarize_sleep.R")
output_path <- file.path(job_dir, "sleep-summary.csv")

writeLines(
  c(
    "result <- data.frame(",
    "  mean_extra = mean(datasets::sleep$extra),",
    "  n = nrow(datasets::sleep)",
    ")",
    sprintf("write.csv(result, %s, row.names = FALSE)", shQuote(output_path))
  ),
  con = script_path
)
```

## Build a Job

Start with `rjob()` and add properties with `+`. This first job uses the local
backend so you can run it before connecting to a cluster. Property functions
are named for the part of the job they update:

- `script()` stores the script path.
- `job_directory()` stores the working directory for the job.
- `scheduler()` chooses the execution backend.
- `resources()` records scheduler resource requests.
- `packages()` declares packages to load for the job.
- `libraries()` explicitly configures R library search paths when needed.

```{r}
sleep_job <- rjob("sleep_summary") +
  script(script_path) +
  job_directory(job_dir) +
  scheduler("local") +
  resources(
    n_nodes = 1,
    n_cores = 1,
    wall_time = format_wall_time(minutes = 10),
    total_memory = "1G"
  ) +
  packages("stats")

sleep_job
```

The object stores the job specification. It does not submit anything when you
create or print it.

```{r}
summary(sleep_job)
```

## Packages and R Libraries

`packages()` declares packages a job needs. At submission, `hpcR` checks the
submission session's effective library search path and passes that path to the
job's R process through `R_LIBS`. The runner then verifies that every declared
package is available before attaching any of them.

By default, missing packages stop submission with a clear error. Installation
is never attempted on a compute node. To have the submission session offer to
install missing packages, set `install = "ask"`; for non-interactive
submission workflows, use `install = "always"`. Both choices install into one
writable library, defaulting to `R_LIBS_USER` when it identifies a single path.

For example, this job asks before installing missing packages in a project
library:

```{r, eval = FALSE}
analysis_job <- rjob("analysis") +
  script("/path/to/analysis.R") +
  job_directory("/path/to/project") +
  scheduler("slurm") +
  packages(
    c("data.table", "targets"),
    install = "ask",
    install_library = "/shared/project/R/library"
  )
```

Use `libraries()` when you need to make an additional library available to
the submission and compute environments: for example, your cluster requires a
site library, or a script uses `pkg::function()` without listing that package
for attachment. Explicit `R_LIBS` paths are searched before paths supplied by
the submission session's existing R library configuration.

```{r, eval = FALSE}
analysis_job <- analysis_job +
  libraries(
    R_LIBS = "/shared/site/R/library",
    R_LIBS_USER = "~/R/library"
  )
```

Both kinds of paths must be visible from the compute node and compatible with
the R installation used there. Prefer `packages()` for packages that the job
attaches and pre-submission installation policy; use `libraries()` for
exceptional or broader R-library setup.

## Run Your First Job Locally

On a UNIX-like system, run the job locally to confirm that the script, working
directory, and package setup are correct. This starts a background process,
waits for it to finish, and reads the CSV file written by the script.

```{r, eval = FALSE}
local_job_id <- sleep_job |> submit()

wait_for_job(
  job_ids = local_job_id,
  scheduler_name = "local",
  repolling_interval = 0.1,
  max_wait = 30
)

read.csv(output_path)
```

The final command should show one row with the mean extra sleep and the number
of observations. On Windows, skip this local step and use the cluster backend
available in your environment.

## Choose a Scheduler

`hpcR` currently recognizes Slurm, TORQUE, and a local shell backend:

```{r, eval = FALSE}
scheduler("slurm")   # alias: scheduler("sbatch")
scheduler("torque")  # alias: scheduler("qsub")
scheduler("local")   # alias: scheduler("sh")
```

Use the scheduler name that matches the environment where the R session is
submitting the job. After the local check succeeds, change the backend for a
cluster submission:

```{r, eval = FALSE}
cluster_job <- sleep_job + scheduler("slurm")
```

## Submit a Job

Use `submit()` for R scripts configured with `rjob()`. `hpcR` then runs the
script with the appropriate R job wrapper on the selected backend.

```{r, eval = FALSE}
job_id <- cluster_job |> submit()
job_id
```

On a Slurm system this calls `sbatch`; on a TORQUE system it calls `qsub`; with
the local backend it starts the process locally and returns the process ID.

## Check and Wait for Jobs

After submission, pass the returned job ID to `check_job_status()`:

```{r, eval = FALSE}
check_job_status(
  job_ids = job_id,
  scheduler_name = "slurm"
)
```

To pause an R workflow until one or more jobs finish, use `wait_for_job()`:

```{r, eval = FALSE}
wait_for_job(
  job_ids = job_id,
  scheduler_name = "slurm",
  repolling_interval = 60,
  max_wait = 60 * 60
)
```

`wait_for_job()` is useful in pipeline scripts where a parent R process submits
work and then needs to wait before continuing.

## Submit Dependent Jobs

For scheduler-level dependencies, use `depends_on` in `submit_job()`. The
dependency IDs are passed through to the scheduler-specific submission command.

```{r, eval = FALSE}
first_id <- submit_job(
  input = "prepare_data.R",
  input_type = "script",
  scheduler_name = "slurm"
)

second_id <- submit_job(
  input = "fit_model.R",
  input_type = "script",
  scheduler_name = "slurm",
  depends_on = first_id
)
```

For local jobs, `depends_on` causes `hpcR` to wait for the dependent process IDs
before starting the next local job.

## Practical Workflow

A typical script that submits work to a cluster looks like this:

```{r, eval = FALSE}
library(hpcR)

analysis_job <- rjob("analysis") +
  script("/path/to/analysis.R") +
  job_directory("/path/to/project") +
  scheduler("slurm") +
  resources(
    n_nodes = 1,
    n_cores = 4,
    wall_time = format_wall_time(hours = 2),
    total_memory = "16G"
  ) +
  packages(c("data.table", "stats"))

summary(analysis_job)

job_id <- submit(analysis_job)
check_job_status(job_id, scheduler_name = "slurm")
```

The most important habit is to build and inspect the job before submitting it.
That keeps the scheduler settings, script path, working directory, and package
requirements visible in one place. For preflight checks, compiled submission
metadata, and troubleshooting, see
`vignette("inspecting-and-debugging-submissions", package = "hpcR")`.
